Modelling BCG vaccination in the UK

### 2019-07-04 rmds/06-tb-bcg-improv-outcomes.Rmd
checked accroynm usage

To conduct this study, I first obtained all TB notifications for 2009-2015 in England from the Enhanced Tuberculosis surveillance system (see Chapter \@ref(data)). I then considered five outcomes: All-cause mortality, death due to TB (in those who died), recurrent TB, pulmonary disease, and sputum smear status. I used logistic regression, with complete case analysis, to investigate each outcome with BCG vaccination, years since vaccination and age at vaccination, adjusting for potential confounders. All analyses were repeated using multiply imputed data. This work was supervised by Hannah Christensen and Ellen Brooks-Pollock. Collaborators at Public Health England including, Maeve K Lalor, Dominik Zenner, Colin Campbell, and Mary E Ramsay provided the data and commented on multiple versions of this work.

@@

Bacillus Calmette–Guérin (BCG) is one of the mostly widely-used vaccines and the only vaccine that protects against tuberculosis (TB) disease. BCG was first used in humans in 1921 and was introduced into the WHO Expanded Program on Immunization in 1974.[@The2004] BCG vaccination has been controversial due to its variable efficacy and possibility of causing a false positive result with the standard skin test for TB.[@Zwerling2011a] However, the lack of a more effective vaccine and the emergence of drug-resistant TB strains means that BCG vaccination remains an important tool for reducing TB incidence and mortality rates.

@@

### Enhanced Tuberculosis Surveillance (ETS) system

I extracted all notifications from the Enhanced Tuberculosis Surveillance (ETS) system from January 1, 2009 to December 31, 2015 (Chapter \@ref(data)). BCG vaccination status and year of vaccination have been collected since 2008. The outcomes I considered were: all-cause mortality, death due to TB (in those who died), recurrent TB, pulmonary disease, and sputum smear status. These outcomes were selected based on: their availability in the ETS; evidence from the literature of prior associations with BCG vaccination; associations with increased case infectiousness; or severe outcomes for patients.

@@

outcome_demo_tab_bcg %>% 
  set_names(., c("Outcome", names(.)[-1])) %>% 
  kable(caption = "Outcomes for individuals in England notified with tuberculosis between 2009-2015,  stratified by BCG vaccination status.",
         booktabs = TRUE, escape = TRUE) %>% 
    column_spec(c(1), width = "2cm") %>% 
    add_indent(indents) %>% 
    kable_styling(font_size = 8) %>% 
    kable_styling(latex_options = "hold_position")%>% 
    add_header_above(c(" " = 2, "BCG status" = 3)) %>% 
    add_footnote(label = c("{% all cases}(% complete within vaccine status)[% complete within category]",
 "* Death due to TB in those who died and where cause of death was known"
), notation = "none") %>% 
  {.}
```

@@

demo_tab_bcg %>% 
  set_names(., c("Confounder", names(.)[-1])) %>% 
  kable(caption = "Confounders for individuals in England notified with tuberculosis between 2009-2015,  stratified by BCG vaccination status.",
         booktabs = TRUE, escape = TRUE) %>% 
    column_spec(c(1), width = "2cm") %>% 
    add_indent(indents) %>% 
    kable_styling(font_size = 8) %>% 
    kable_styling(latex_options = "hold_position") %>% 
    add_header_above(c(" " = 2, "BCG status" = 3)) %>% 
    add_footnote(label = c("{% all cases}(% complete within vaccine status)[% complete within category]",
 "* Death due to TB in those who died and where cause of death was known"
), notation = "none") %>% 
  {.}
```

>>>>

To conduct this study, I first obtained all TB notifications for 2009-2015 in England from the Enhanced Tuberculosis surveillance (ETS) system (see Chapter \@ref(data)). I then considered five outcomes: All-cause mortality, death due to TB (in those who died), recurrent TB, pulmonary disease, and sputum smear status. I used logistic regression, with complete case analysis, to investigate each outcome with BCG vaccination, years since vaccination and age at vaccination, adjusting for potential confounders. All analyses were repeated using multiply imputed data. This work was supervised by Hannah Christensen and Ellen Brooks-Pollock. Collaborators at Public Health England including, Maeve K Lalor, Dominik Zenner, Colin Campbell, and Mary E Ramsay provided the data and commented on multiple versions of this work.

@@

Bacillus Calmette–Guérin (BCG) is one of the mostly widely-used vaccines and the only vaccine that protects against TB disease. BCG was first used in humans in 1921 and was introduced into the WHO Expanded Program on Immunization in 1974.[@The2004] BCG vaccination has been controversial due to its variable efficacy and possibility of causing a false positive result with the standard skin test for TB.[@Zwerling2011a] However, the lack of a more effective vaccine and the emergence of drug-resistant TB strains means that BCG vaccination remains an important tool for reducing TB incidence and mortality rates.

@@

### Enhanced TB Surveillance (ETS) system

I extracted all notifications from the ETS system from January 1, 2009 to December 31, 2015 (Chapter \@ref(data)). BCG vaccination status and year of vaccination have been collected since 2008. The outcomes I considered were: all-cause mortality, death due to TB (in those who died), recurrent TB, pulmonary disease, and sputum smear status. These outcomes were selected based on: their availability in the ETS; evidence from the literature of prior associations with BCG vaccination; associations with increased case infectiousness; or severe outcomes for patients.

@@

outcome_demo_tab_bcg %>% 
  set_names(., c("Outcome", names(.)[-1])) %>% 
  kable(caption = "Outcomes for individuals in England notified with TB between 2009-2015,  stratified by BCG vaccination status.",
         booktabs = TRUE, escape = TRUE) %>% 
    column_spec(c(1), width = "2cm") %>% 
    add_indent(indents) %>% 
    kable_styling(font_size = 8) %>% 
    kable_styling(latex_options = "hold_position")%>% 
    add_header_above(c(" " = 2, "BCG status" = 3)) %>% 
    add_footnote(label = c("{% all cases}(% complete within vaccine status)[% complete within category]",
 "* Death due to TB in those who died and where cause of death was known"
), notation = "none") %>% 
  {.}
```

@@

demo_tab_bcg %>% 
  set_names(., c("Confounder", names(.)[-1])) %>% 
  kable(caption = "Confounders for individuals in England notified with TB between 2009-2015,  stratified by BCG vaccination status.",
         booktabs = TRUE, escape = TRUE) %>% 
    column_spec(c(1), width = "2cm") %>% 
    add_indent(indents) %>% 
    kable_styling(font_size = 8) %>% 
    kable_styling(latex_options = "hold_position") %>% 
    add_header_above(c(" " = 2, "BCG status" = 3)) %>% 
    add_footnote(label = c("{% all cases}(% complete within vaccine status)[% complete within category]",
 "* Death due to TB in those who died and where cause of death was known"
), notation = "none") %>% 
  {.}
```

### 2019-07-04 rmds/07-direct-effect-bcg-change.Rmd
checked accroynm usage

I combined notification data from the Enhanced Tuberculosis Surveillance system, with demographic data from the Labour Force Survey to construct retrospective cohorts of individuals in England relevant to both the universal, and targeted vaccination programmes between Jan 1, 2000 and Dec 31, 2010. For each cohort, I estimated incidence rates over a 5 year follow-up period and used Poisson and Negative Binomial regression models in order to estimate the impact of the change in policy on TB. This work was supervised by Hannah Christensen and Ellen Brooks-Pollock. Nicky Weltton provided guidance on the statistical methods used.

@@

In 2005 England changed its Bacillus Calmette–Guérin (BCG) vaccination policy against tuberculosis (TB) from a universal programme aimed at 13 and 14 year olds to a targeted programme aimed at high-risk neonates (see Chapter \@ref(background)). High risk babies are identified by local TB incidence and by the parents’ and grandparents’ country of origin. The change in policy was motivated by evidence of reduced TB transmission,[@PHE2016a; @Fine2005a; @Teo2006] and high effectiveness of the BCG vaccine in children,[@Rodrigues1993; @Colditz1994; @Roy2014] and variable effectiveness in adults.[Mangtani2014; @Zwerling2011] Little work has been done to evaluate the impact of this change in vaccination policy.

@@

Data on all notifications from the Enhanced Tuberculosis Surveillance (ETS) system from Jan 1, 2000 to Dec 31, 2015 were obtained from Public Health England (PHE). The ETS is maintained by PHE, and contains demographic, clinical, and microbiological data on all notified cases in England (see Chapter \@ref(data)). A descriptive analysis of TB epidemiology in England is published each year, which fully details data collection and cleaning.[@PHE2016a]

@@

As I was imputing a single variable I reformulated the imputation as a categorical prediction problem. This allowed the use of techniques from machine learning to improve the quality of our imputation, whilst also validating it using metrics supported by theory. I included year of notification, sex, age, Public Health England Centre (PHEC), occupation, ethnic group, Index of Multiple Deprivation (2010) categorised into five groups for England (IMD rank), and risk factor count (risk factors considered; drug use, homelessness, alcohol misuse/abuse and prison). However, I could not account for a possible missing not at random mechanism not captured by these covariates. To train the model I first split the data with complete UK birth status into a training set (80%), a calibration set (5%), and a test set (15%). I then fit a gradient boosted machine with 10,000 trees, early stopping (at a precision of $1e-5$, with 10 stopping rounds), a learning rate of 0.1, and a learn rate annealing of 0.99. Gradient boosted machines are a tree based method that can incorporate complex non-linear relationships and interactions. Much like a random forest model they work by ensembling a group of trees, but unlike a random forest model each tree is additive aiming to reduce the residual loss from previous trees. Once the model had been fit to the training set I performed platt scaling using the calibration data set. Our fitted imputation model had a Logloss of 0.28 on the test set, with an AUC of 0.93, both of which indicate a robust out of bag performance. I found that ethnic group was the most important variable for predicting UK birth status, followed by age and PHEC.

@@

I was unable to estimate the impact of the change in BCG policy after 5 years post vaccination, so both the estimates of the positive and negative consequences are likely to be underestimates of the ongoing impact. Tuberculosis is a complex disease and the BCG vaccine is known to offer imperfect protection, which has been shown to vary both spatially and with time since vaccination (see Chapter \@ref(background)).[@Mangtani2014a; @Abubakar2013] By focusing on the impact of the change in policy on the directly affected populations within a short period of time, and by employing a multi-model approach I have limited the potential impact of these issues. This study was based on a routine observational data set (ETS), and a repeated survey (LFS) both of which may have introduced bias. Whilst the LFS is a robust data source, widely used in academic studies,[@French2007; @Davies2016a; @Lindley2009] it is susceptible to sampling errors particularly in the young, and in the old, which may have biased the estimated incidence rates. As the ETS is routine surveillance system some level of missing data is inevitable (see Chapter \@ref(data)). However, UK birth status is relatively complete (93% (106765/114820)) and I imputed missing values using an approach which accounted for MNAR mechanisms for the variables included in the imputation model. I was unable to adjust for known demographic risk factors for TB, notably socio-economic status,[@Parslow2001; @Bhatti1995] and ethnicity.[@Parslow2001; @Bhatti1995; @Abubakar2008] However, this confounding is likely to be mitigated by the use of multiple cohorts and the adjustment for incidence rates in the UK born and non-UK born. Finally, I have assumed that the effect I have estimated for the change in BCG policy is due to the changes in BCG vaccination policy as well as other associated changes in TB control policy, after adjusting for hypothesised confounders. However, there may have been additional policy changes which I have not accounted for.

>>>>

I combined notification data from the Enhanced TB Surveillance (ETS) system, with demographic data from the Labour Force Survey to construct retrospective cohorts of individuals in England relevant to both the universal, and targeted vaccination programmes between Jan 1, 2000 and Dec 31, 2010. For each cohort, I estimated incidence rates over a 5 year follow-up period and used Poisson and Negative Binomial regression models in order to estimate the impact of the change in policy on TB. This work was supervised by Hannah Christensen and Ellen Brooks-Pollock. Nicky Weltton provided guidance on the statistical methods used.

@@

In 2005 England changed its Bacillus Calmette–Guérin (BCG) vaccination policy against TB (TB) from a universal programme aimed at 13 and 14 year olds to a targeted programme aimed at high-risk neonates (see Chapter \@ref(background)). High risk babies are identified by local TB incidence and by the parents’ and grandparents’ country of origin. The change in policy was motivated by evidence of reduced TB transmission,[@PHE2016a; @Fine2005a; @Teo2006] and high effectiveness of the BCG vaccine in children,[@Rodrigues1993; @Colditz1994; @Roy2014] and variable effectiveness in adults.[Mangtani2014; @Zwerling2011] Little work has been done to evaluate the impact of this change in vaccination policy.

@@

Data on all notifications from the ETS system from Jan 1, 2000 to Dec 31, 2015 were obtained from Public Health England (PHE). The ETS is maintained by PHE, and contains demographic, clinical, and microbiological data on all notified cases in England (see Chapter \@ref(data)). A descriptive analysis of TB epidemiology in England is published each year, which fully details data collection and cleaning.[@PHE2016a]

@@

As I was imputing a single variable I reformulated the imputation as a categorical prediction problem. This allowed the use of techniques from machine learning to improve the quality of our imputation, whilst also validating it using metrics supported by theory. I included year of notification, sex, age, PHE Centre (PHEC), occupation, ethnic group, Index of Multiple Deprivation (2010) categorised into five groups for England (IMD rank), and risk factor count (risk factors considered; drug use, homelessness, alcohol misuse/abuse and prison). However, I could not account for a possible missing not at random mechanism not captured by these covariates. To train the model I first split the data with complete UK birth status into a training set (80%), a calibration set (5%), and a test set (15%). I then fit a gradient boosted machine with 10,000 trees, early stopping (at a precision of $1e-5$, with 10 stopping rounds), a learning rate of 0.1, and a learn rate annealing of 0.99. Gradient boosted machines are a tree based method that can incorporate complex non-linear relationships and interactions. Much like a random forest model they work by ensembling a group of trees, but unlike a random forest model each tree is additive aiming to reduce the residual loss from previous trees. Once the model had been fit to the training set I performed platt scaling using the calibration data set. Our fitted imputation model had a Logloss of 0.28 on the test set, with an AUC of 0.93, both of which indicate a robust out of bag performance. I found that ethnic group was the most important variable for predicting UK birth status, followed by age and PHEC.

@@

I was unable to estimate the impact of the change in BCG policy after 5 years post vaccination, so both the estimates of the positive and negative consequences are likely to be underestimates of the ongoing impact. TB is a complex disease and the BCG vaccine is known to offer imperfect protection, which has been shown to vary both spatially and with time since vaccination (see Chapter \@ref(background)).[@Mangtani2014a; @Abubakar2013] By focusing on the impact of the change in policy on the directly affected populations within a short period of time, and by employing a multi-model approach I have limited the potential impact of these issues. This study was based on a routine observational data set (ETS), and a repeated survey (LFS) both of which may have introduced bias. Whilst the LFS is a robust data source, widely used in academic studies,[@French2007; @Davies2016a; @Lindley2009] it is susceptible to sampling errors particularly in the young, and in the old, which may have biased the estimated incidence rates. As the ETS is routine surveillance system some level of missing data is inevitable (see Chapter \@ref(data)). However, UK birth status is relatively complete (93% (106765/114820)) and I imputed missing values using an approach which accounted for MNAR mechanisms for the variables included in the imputation model. I was unable to adjust for known demographic risk factors for TB, notably socio-economic status,[@Parslow2001; @Bhatti1995] and ethnicity.[@Parslow2001; @Bhatti1995; @Abubakar2008] However, this confounding is likely to be mitigated by the use of multiple cohorts and the adjustment for incidence rates in the UK born and non-UK born. Finally, I have assumed that the effect I have estimated for the change in BCG policy is due to the changes in BCG vaccination policy as well as other associated changes in TB control policy, after adjusting for hypothesised confounders. However, there may have been additional policy changes which I have not accounted for.

### 2019-07-04 rmds/08-model-development.Rmd
checked accroynm usage

This chapter presents the development and parameterisation of a, semi-stochastic, dynamic model of TB transmission, incorporating BCG vaccination, in the UK born population of England. The key features of TB transmission, and BCG vaccination, are discussed with details of pertinent TB models given. An appropriate model structure for answering the study question is then outlined, along with a justification of the choices made and details of required sensitivity analyses. The model structure is then defined mathematically and parameterised using literature sources as well as data from the Enhanced Tuberculosis Surveillance System (ETS), Labour Force Survey (LFS) and Office for National Statistics (ONS) (see Chapter \@ref(data)). The assumptions made during model building and parameterisation are highlighted in preparation for evaluation during model fitting (Chapter \@ref(model-fitting)).

@@

When developing an infectious disease dynamic model there is a trade-off between reproducing reality and interpretability.[@Anderson1991] A model that includes all known features of a disease may not be able to answer questions of interest as it is too complex to interpret or because data does not exist to calibrate many of its parameters. A highly complex model, or indeed an overly simplistic one, may also be at risk of bias. The optimal model is therefore as parsimonious as possible, whilst still capturing the key features of a disease and making best use of all available data.[@Anderson1991] In this section the key features of TB, and BCG vaccination, that must be captured in order to produce meaningful output are discussed, as well as the features that can be excluded for this study question. Data from the Enhanced Tuberculosis Surveillance (ETS) (Chapter \@ref(data)) is used to support evidence from the literature. Further background information can be found in Chapter \@ref(background) and Chapter \@ref(data).

## Tuberculosis disease

@@

```{r model-flow-diag, out.width = "400px", fig.cap = "Flow diagram for the dynamic Tuberculosis (TB) disease model with demographics and vaccination described in the text. The TB model contains the following compartments; Susceptible ($S$), high risk latent ($H$), low risk latent ($L$), active cases with pulmonary TB ($P$), active TB cases with extra-pulmonary TB only ($E$), pulmonary cases on treatment ($T_P$), and extra-pulmonary cases on treatment ($T_E$). The vaccinated ($v$) and unvaccinated ($u$) populations are represented by $k$, such that $k = u,v$. Age stratification is represented by $a$ (where $a = 1, 2, ...,11$) in the disease model and the $0,\ 1,\ 2,\ 3$ subscripts in the demographic model. Five year age groups are modelled (i.e $0-4$, $5-9$, $10-14$, ...) up to 49 years old, with a single age group for those aged 50-69 years old and those aged 70-89 years old. Individuals aged $90+$ are not explicitly modelled. In the demographic and vaccination model the A compartment represents the demographic processes modelled in all population compartments except for the vacinated and unvaccinated susceptible populations. Solid arrows represent transition rates within the modelled populations and dashed arrows represent transition rates into, or out of the modelled populations (i.e birth and death processes)."}
knitr::include_graphics(file.path(resources_path, "resources", "model_diagrams", "tb_model_flow.png"))
```

@@

Parameters distributions were either estimated from the available data or assumed based on common values found in the literature. Where no comparable estimates were found in the literature a largely uninformative distribution has been used. The discussion of model parameters has been split into disease model parameters, vaccination model parameters, and demographic model parameters. The data sources used to estimate model parameters have been detailed although for the Enhanced TB Surveillance system (ETS) and the Labour Force Survey more detail is provided elsewhere (Chapter \@ref(data)).

@@

```{r tb-time-treat, fig.cap = "Distribution of time to treatment (days) from the date of reported symptom onset until the date started treatment for the UK born, stratified by age group and pulmonary/extra-pulmonary TB status in the Enhanced Tuberculosis Surveillance system for notifications between 2000 and 2012. Age is stratified into three groups; children (0-14), adults (15-59) older adults (15-59). The time from symptom onset to starting treatment is shorter for cases with pulmonary TB cases across age groups, with younger cases starting treatment more rapidly than older cases."}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "time_to_treatment.png"))
```

@@

```{r tb-treat-time-succ, fig.cap = "Distribution of time to treatment completion in the UK born successfully treated (days), stratified by age group and pulmonary/extra-pulmonary TB status in the Enhanced Tuberculosis Surveillance system for notifications between 2000 and 2012. Age is stratified into three groups; children (0-14), adults (15-59) older adults (15-59). There is little evidence that the time to successful treatment differs between pulmonary and extra-pulmonary cases only but older cases appear to have a high likelihood of longer treatment times."}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "time_cure.png"))
```

@@

Age-stratified and UK birth stratified population estimates for England were estimated using the Labour Force Survey (Chapter \@ref(data)). Figure \@ref(fig:age-strat-demo) indicates that the age distribution of the UK born population changed over the study period (2000 to 2015), with an increase in those in late middle age and older and a decrease in those in early middle age. The proportion of young adults and young children also increased. This may have impacted TB incidence as young adults are thought to be responsible for the majority of transmission. Data from the 1931 census was also used to estimate the population of England in 1931 stratified into the modeled age groups.

@@

Several studies have made use of dynamic TB transmission models to evaluate BCG vaccination or future vaccines.[@Harris2016; @Egbetade2011a; @Bhunu2008b] In general, these studies used less detailed models than the one presented here, typically because they were modelling TB in a more generic setting or because more information about TB epidemiology, TB natural history, and the BCG vaccine has become available over time. In addition, no dynamic model of TB, including BCG vaccination, has currently been published that includes both protection from initial infection and protection from active TB due to BCG vaccination. There have also been no studies that use the current best estimates for the duration of BCG protection in developed countries away from the equator. Harris et al. reviewed mathematical models that explored the epidemiological impacts of future TB vaccines. They found that vaccines targeted at all-ages or at adolescents/adults were more effective at eradicating TB than neonatal programmes when vaccine effectiveness was not assumed to degrade with age. The majority of studies included in their review used deterministic, compartmental, dynamic models. Model structures were found to have evolved over time as TB natural history and epidemiology is better understood, with the majority of models having at least susceptible, latent, active disease, and recovered states. Treatment status, variable infectiousness of active disease, vaccine waning, and age stratification were included in some of the models evaluated.[@Harris2016] Recently it has been shown that only models that include at least two latent compartments are able to reproduce the observed activation dynamics of TB.[@Ragonnet2017] The model presented here is based on the serial latency archetype identified in this study. It has also been shown that realistic age structure and population demographics, included in the model presented in this chapter, are critical for reproducing TB epidemiology.[@Brooks-Pollock2010] Egbetade et al. presented a dynamic model of TB that included BCG vaccination but did not include age structure. They found that universal vaccination increased the stability of the disease free equilibrium in countries with high TB burden. However the model presented was not rigorously parameterised with data and only a single latent TB compartment was used.[@Egbetade2011a] Bhunu et al. developed a dynamic transmission model of TB that in order to investigate the effects of pre- and post-exposure vaccines for tuberculosis control. Again their model did not include multiple latent compartments or age structure unlike the model presented here.[@Bhunu2008b]

>>>>

This chapter presents the development and parameterisation of a, semi-stochastic, dynamic model of TB transmission, incorporating BCG vaccination, in the UK born population of England. The key features of TB transmission, and BCG vaccination, are discussed with details of pertinent TB models given. An appropriate model structure for answering the study question is then outlined, along with a justification of the choices made and details of required sensitivity analyses. The model structure is then defined mathematically and parameterised using literature sources as well as data from the Enhanced TB Surveillance System (ETS), Labour Force Survey (LFS) and Office for National Statistics (ONS) (see Chapter \@ref(data)). The assumptions made during model building and parameterisation are highlighted in preparation for evaluation during model fitting (Chapter \@ref(model-fitting)).

@@

When developing an infectious disease dynamic model there is a trade-off between reproducing reality and interpretability.[@Anderson1991] A model that includes all known features of a disease may not be able to answer questions of interest as it is too complex to interpret or because data does not exist to calibrate many of its parameters. A highly complex model, or indeed an overly simplistic one, may also be at risk of bias. The optimal model is therefore as parsimonious as possible, whilst still capturing the key features of a disease and making best use of all available data.[@Anderson1991] In this section the key features of TB, and BCG vaccination, that must be captured in order to produce meaningful output are discussed, as well as the features that can be excluded for this study question. Data from the ETS (Chapter \@ref(data)) is used to support evidence from the literature. Further background information can be found in Chapter \@ref(background) and Chapter \@ref(data).

## TB disease

@@

```{r model-flow-diag, out.width = "400px", fig.cap = "Flow diagram for the dynamic TB disease model with demographics and vaccination described in the text. The TB model contains the following compartments; Susceptible ($S$), high risk latent ($H$), low risk latent ($L$), active cases with pulmonary TB ($P$), active TB cases with extra-pulmonary TB only ($E$), pulmonary cases on treatment ($T_P$), and extra-pulmonary cases on treatment ($T_E$). The vaccinated ($v$) and unvaccinated ($u$) populations are represented by $k$, such that $k = u,v$. Age stratification is represented by $a$ (where $a = 1, 2, ...,11$) in the disease model and the $0,\ 1,\ 2,\ 3$ subscripts in the demographic model. Five year age groups are modelled (i.e $0-4$, $5-9$, $10-14$, ...) up to 49 years old, with a single age group for those aged 50-69 years old and those aged 70-89 years old. Individuals aged $90+$ are not explicitly modelled. In the demographic and vaccination model the A compartment represents the demographic processes modelled in all population compartments except for the vacinated and unvaccinated susceptible populations. Solid arrows represent transition rates within the modelled populations and dashed arrows represent transition rates into, or out of the modelled populations (i.e birth and death processes)."}
knitr::include_graphics(file.path(resources_path, "resources", "model_diagrams", "tb_model_flow.png"))
```

@@

Parameters distributions were either estimated from the available data or assumed based on common values found in the literature. Where no comparable estimates were found in the literature a largely uninformative distribution has been used. The discussion of model parameters has been split into disease model parameters, vaccination model parameters, and demographic model parameters. The data sources used to estimate model parameters have been detailed although for the ETS and the LFS more detail is provided elsewhere (Chapter \@ref(data)).

@@

```{r tb-time-treat, fig.cap = "Distribution of time to treatment (days) from the date of reported symptom onset until the date started treatment for the UK born, stratified by age group and pulmonary/extra-pulmonary TB status in the ETS system for notifications between 2000 and 2012. Age is stratified into three groups; children (0-14), adults (15-59) older adults (15-59). The time from symptom onset to starting treatment is shorter for cases with pulmonary TB cases across age groups, with younger cases starting treatment more rapidly than older cases."}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "time_to_treatment.png"))
```

@@

```{r tb-treat-time-succ, fig.cap = "Distribution of time to treatment completion in the UK born successfully treated (days), stratified by age group and pulmonary/extra-pulmonary TB status in the ETS for notifications between 2000 and 2012. Age is stratified into three groups; children (0-14), adults (15-59) older adults (15-59). There is little evidence that the time to successful treatment differs between pulmonary and extra-pulmonary cases only but older cases appear to have a high likelihood of longer treatment times."}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "time_cure.png"))
```

@@

Age-stratified and UK birth stratified population estimates for England were estimated using the LFS (Chapter \@ref(data)). Figure \@ref(fig:age-strat-demo) indicates that the age distribution of the UK born population changed over the study period (2000 to 2015), with an increase in those in late middle age and older and a decrease in those in early middle age. The proportion of young adults and young children also increased. This may have impacted TB incidence as young adults are thought to be responsible for the majority of transmission. Data from the 1931 census was also used to estimate the population of England in 1931 stratified into the modeled age groups.

@@

Several studies have made use of dynamic TB transmission models to evaluate BCG vaccination or future vaccines.[@Harris2016; @Egbetade2011a; @Bhunu2008b] In general, these studies used less detailed models than the one presented here, typically because they were modelling TB in a more generic setting or because more information about TB epidemiology, TB natural history, and the BCG vaccine has become available over time. In addition, no dynamic model of TB, including BCG vaccination, has currently been published that includes both protection from initial infection and protection from active TB due to BCG vaccination. There have also been no studies that use the current best estimates for the duration of BCG protection in developed countries away from the equator. Harris et al. reviewed mathematical models that explored the epidemiological impacts of future TB vaccines. They found that vaccines targeted at all-ages or at adolescents/adults were more effective at eradicating TB than neonatal programmes when vaccine effectiveness was not assumed to degrade with age. The majority of studies included in their review used deterministic, compartmental, dynamic models. Model structures were found to have evolved over time as TB natural history and epidemiology is better understood, with the majority of models having at least susceptible, latent, active disease, and recovered states. Treatment status, variable infectiousness of active disease, vaccine waning, and age stratification were included in some of the models evaluated.[@Harris2016] Recently it has been shown that only models that include at least two latent compartments are able to reproduce the observed activation dynamics of TB.[@Ragonnet2017] The model presented here is based on the serial latency archetype identified in this study. It has also been shown that realistic age structure and population demographics, included in the model presented in this chapter, are critical for reproducing TB epidemiology.[@Brooks-Pollock2010] Egbetade et al. presented a dynamic model of TB that included BCG vaccination but did not include age structure. They found that universal vaccination increased the stability of the disease free equilibrium in countries with high TB burden. However the model presented was not rigorously parameterised with data and only a single latent TB compartment was used.[@Egbetade2011a] Bhunu et al. developed a dynamic transmission model of TB that in order to investigate the effects of pre- and post-exposure vaccines for TB control. Again their model did not include multiple latent compartments or age structure unlike the model presented here.[@Bhunu2008b]

### 2019-07-04 rmds/09-model-fitting.Rmd
checked accroynm usage

The primary data source used was the reported UK born TB notifications from 2000 to 2004 as recorded by the Enhanced Tuberculosis Survelliance (ETS) system (see Chapter \@ref(data)). These are the only years for which notifications are available stratified by UK birth status and for which universal school-age BCG vaccination was in place. As the model of TB transmission outlined in the previous chapter had multiple age-dependent parameters UK born TB notifications stratified by age group are also fitted to. The age groups considered are children (aged 0-14 years old), adults (15-69 years old), and older adults (70-89 years old). These age groups were used, rather than the more detailed 5 year age groups used in the TB transmission model (see Chapter \@ref(model-development)), for three reasons. Firstly, by using condensed age groups the number of notifications in each group increased. Having a larger number of notifications in each group reduces the impact of stochastics noise, which makes fitting the model easier. Secondly, reducing the number of data points, whilst still capturing the important age dynamics, reduces the compute requirements of the model (see Section \@ref(particle-filter)). As will be discussed in Section \@ref(fitting-pipeline) this was a major consideration as the model fitting approached used was highly compute intensive. Finally, the condensed age groups used were chosen to be meaningful. This allows model results to be more easily reported and interpreted.

>>>>

The primary data source used was the reported UK born TB notifications from 2000 to 2004 as recorded by the Enhanced TB Survelliance (ETS) system (see Chapter \@ref(data)). These are the only years for which notifications are available stratified by UK birth status and for which universal school-age BCG vaccination was in place. As the model of TB transmission outlined in the previous chapter had multiple age-dependent parameters UK born TB notifications stratified by age group are also fitted to. The age groups considered are children (aged 0-14 years old), adults (15-69 years old), and older adults (70-89 years old). These age groups were used, rather than the more detailed 5 year age groups used in the TB transmission model (see Chapter \@ref(model-development)), for three reasons. Firstly, by using condensed age groups the number of notifications in each group increased. Having a larger number of notifications in each group reduces the impact of stochastics noise, which makes fitting the model easier. Secondly, reducing the number of data points, whilst still capturing the important age dynamics, reduces the compute requirements of the model (see Section \@ref(particle-filter)). As will be discussed in Section \@ref(fitting-pipeline) this was a major consideration as the model fitting approached used was highly compute intensive. Finally, the condensed age groups used were chosen to be meaningful. This allows model results to be more easily reported and interpreted.
